Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-15 opencl/papers/foropencl.tex
Added shallow-water.tex
cleaning up verbatim mess

\usepackage{cite}
\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{array}
\usepackage{url}

@@

\input{intro}
\input{programming-model}
\input{source-to-source}
\input{static-analysis}

>>>>

\usepackage{cite}
\usepackage{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{array}
\usepackage{url}
\usepackage{listings}

\lstdefinestyle{FortranLike}{float,frame=lines,language=Fortran,commentstyle=\ttfamily,basicstyle=\ttfamily}
\lstdefinestyle{CLike}{float,frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}
\lstdefinestyle{NoFloatCLike}{frame=lines,language=C,commentstyle=\ttfamily,basicstyle=\ttfamily}

@@

\input{intro}
\input{programming-model}
\input{shallow-water}
\input{source-to-source}
\input{static-analysis}

### 2010-12-15 opencl/papers/programming-model.tex
cleaning up verbatim mess

\begin{verbatim}

 C = A + s*B

\end{verbatim}

@@

\begin{verbatim}
   pure elemental real function elem_example(a, b, s)
      real, intent(in) :: a, b, s
      elem_example = a + s*b
   end function
\end{verbatim}

and called with

\begin{verbatim}
   C = elem_example(A, B, s)
\end{verbatim}

Note that while {\tt elem\_example} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.

@@

\begin{verbatim}

  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx

\end{verbatim}

>>>>

{\small
\begin{verbatim}
 C = A + s*B
\end{verbatim}
}

@@

{\small
\begin{verbatim}
  pure elemental real function foo(a, b, s)
    real, intent(in) :: a, b, s
    foo = a + s*b
  end function
\end{verbatim}
}

\noindent and called with

{\small
\begin{verbatim}
  C = foo(A, B, s)
\end{verbatim}
}

Note that while {\tt foo} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.

@@

{\small
\begin{verbatim}
  Bz = Bz &
     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
\end{verbatim}
}

### 2010-12-15 opencl/papers/shallow-water.tex
cleaning up after Craig....
cleaning up verbatim mess

The Fortran kernel procedure {\tt wave_advance} for the shallow water code
is declared as:
\begin{verbatim}
   pure subroutine wave_advance(H, U, V, dx, dy, dt)
      real, dimension(:,:), intent(inout) :: H, U, V
      real, intent(in) :: dx, dy, dt
      !$OFP CONTIGUOUS :: H, U, V
      !$OFP KERNEL :: wave_advance
   end subroutine
\end{verbatim}
where {\tt H, U}, and {\tt V} are start variables for height and
$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.

Temporary arrays are required for the interior copies {\tt iH, iU}, and
{\tt iV} of the state variables and for the flux quantities {\tt Hx, Hy, Ux,
Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,
\begin{verbatim}
   real, allocatable, dimension(:,:) :: iH, iU, iV
   real, allocatable, dimension(:,:) :: Hx, Hy, Ux, Uy, Vx, Vy
\end{verbatim}

Halos variables for the interior and cell faces are defined as
\begin{verbatim}
   integer, dimension(4) :: halo, face_lt, face_rt, face_up, face_dn
   halo    = [1,1,1,1]
   face_lt = [0,1,1,1]; face_rt = [1,0,1,1]
   face_dn = [1,1,0,1]; face_up = [1,1,1,0]
\end{verbatim}
Note that the halos for the four faces each have a 0 in the definition. Thus the returned
array copy will have size that is larger than the interior regions that use {\tt [1,1,1,1]}.
This is because there is one more cell face quantity than there are cells in a given direction.

The Lax Wendrof (???) update algorithm for the shallow-water model has two steps.
The fluxes are first updated on the cell faces, for example,
\begin{verbatim}
   Hx =           0.5 * ( region(H, face_lt) + region(H, face_rt) ) &
        + (0.5*dt/dx) * ( region(U, face_lt) - region(U, face_rt) )
\end{verbatim}
updates the $x$-direction flux for the height state variable.

Then second step uses these fluxes to update the interior regions for the state varaibles.
For example,
\begin{verbatim}
   face_lt = [0,1,0,0];  face_rt = [1,0,0,0]
   face_dn = [0,0,0,1];  face_up = [0,0,1,0]
   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) ) &
           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) )
\end{verbatim}
Note that face halos have been redefined so that the array copy returned by
now has the same size as the interior region.

>>>>

The Fortran kernel procedure {\tt wave\_advance} for the shallow water code
is declared as:

{\small
\begin{verbatim}
pure subroutine wave_advance(H,U,V,dx,dy,dt)
  real, dimension(:,:), intent(inout) :: H,U,V
  real, intent(in) :: dx,dy,dt
  !$OFP CONTIGUOUS :: H,U,V
  !$OFP KERNEL :: wave_advance
end subroutine
\end{verbatim}
}

where {\tt H, U}, and {\tt V} are start variables for height and
$x$ and $y$ momentum respectively.  The \emph{OFP} compiler directives
{\tt CONTIGUOUS} and {\tt KERNEL} are added because contiguous is Fortran
2008 syntax (not in current compilers) and kernel is an \emph{OFP} extension.

Temporary arrays are required for the interior copies {\tt iH, iU}, and
{\tt iV} of the state variables and for the flux quantities {\tt Hx, Hy, Ux,
Vx}, and {\tt Vy} defined on cell faces.  These temporary arrays are declared as,

{\small
\begin{verbatim}
real, allocatable, dimension(:,:) :: iH,iU,iV
real, allocatable, dimension(:,:) :: Hx,Hy,Ux
real, allocatable, dimension(:,:) :: Uy,Vx,Vy
\end{verbatim}
}

Halos variables for the interior and cell faces are defined as

{\small
\begin{verbatim}
integer, dimension(4) :: halo,face_lt,face_rt
integer, dimension(4) :: face_up,face_dn
halo    = [1,1,1,1]
face_lt = [0,1,1,1]; face_rt = [1,0,1,1]
face_dn = [1,1,0,1]; face_up = [1,1,1,0]
\end{verbatim}
}

Note that the halos for the four faces each have a 0 in the
definition. Thus the returned array copy will have size that is larger
than the interior regions that use {\tt [1,1,1,1]}.  This is because
there is one more cell face quantity than there are cells in a given
direction.

The Lax-Wendroff update algorithm for the shallow-water model has
two steps.  The fluxes are first updated on the cell faces, for
example,

{\small
\begin{verbatim}
Hx = 0.5*(region(H,face_lt)+ &
          region(H,face_rt)) &
   + (0.5*dt/dx) &
   * (region(U,face_lt)-region(U,face_rt))
\end{verbatim}
}

updates the $x$-direction flux for the height state variable.

Then second step uses these fluxes to update the interior regions for
the state variables.  For example,

{\small
\begin{verbatim}
face_lt = [0,1,0,0];  face_rt = [1,0,0,0]
face_dn = [0,0,0,1];  face_up = [0,0,1,0]
iH = iH + (dt/dx) * ( region(Ux, face_lt) - &
                      region(Ux, face_rt) ) &
        + (dt/dy) * ( region(Vy, face_dn) - &
                      region(Vy, face_up) )
\end{verbatim}
}

Note that face halos have been redefined so that the array copy
returned by now has the same size as the interior region.
